Parsing Postfix log files

### 2008-07-09 doc/logparser-acronyms.tex
Corrections from Mads and chktex.

\newacronym{LMTP}{Local Mail Transfer Protocol}{description={
    LMTP is a protocol derived from SMTP that removes the need for the
    server to maintain a mail delivery queue, instead relying on the client
    to maintain it.  Typically the client would be an MTA, and the server
    would be a delivery agent or a mail store.  Full details are available
    in \cite{lmtp-rfc}.
}}

\newacronym{MTA}{Mail Transfer Agent}{description={
    A Mail Transfer Agent sends and/or receives mail via SMTP.
}}

\newacronym{RFC}{Request For Comments}{description={
    The Request For Comments series is a series of proposals defining
    various protocols and file formats, e.g. SMTP.  The name is somewhat
    misleading: initially the authors were asking for peer review, but
    these documents are now the de facto standards the Internet runs on.
}}

\newacronym{SQL}{Structured Query Language}{description={
    SQL is the standard language used for database querying, modification
    and maintenance.  Some information about SQL, including its history,
    can be found at \cite{Wikipedia-sql}; a good introduction can be found
    at~\cite{sql-for-web-nerds}, others are~\cite{w3schools-sql-tutorial,
    sqlcourse.com}.
}}

>>>>

\newacronym{LMTP}{Local Mail Transfer Protocol}{description={
    LMTP is a protocol derived from SMTP that removes the need for the
    server to maintain a mail delivery queue, instead relying on the client
    to maintain it.  Typically the client would be an MTA, and the server
    would be a delivery agent or a mail store.  Full details are available
    in~\cite{lmtp-rfc}.
}}

\newacronym{MTA}{Mail Transfer Agent}{description={
    A Mail Transfer Agent sends and/or receives mail via SMTP\backslashat{}.
}}

\newacronym{RFC}{Request For Comments}{description={
    The Request For Comments series is a series of proposals defining
    various protocols and file formats, e.g.\ SMTP\backslashat{}.  The name
    is somewhat misleading: initially the authors were asking for peer
    review, but these documents are now the de facto standards the Internet
    runs on.
}}

\newacronym{SQL}{Structured Query Language}{description={
    SQL is the standard language used for database querying, modification
    and maintenance.  Some information about SQL, including its history,
    can be found at~\cite{Wikipedia-sql}; a good introduction can be found
    at~\cite{sql-for-web-nerds}, others are~\cite{w3schools-sql-tutorial,
    sqlcourse.com}.
}}

### 2008-07-09 doc/logparser-background.tex
Corrections from Mads and chktex.

XXX CLARIFY THE SENTENCE STARTING WITH "Although this may seem" IN THE NEXT
PARAGRAPH\@.

@@

XXX CHANGE THIS; THE PARSER DESIGN IS NOVEL, DO NOT STRESS THE SIMILARITY
TOO MUCH\@.

@@

There only appears to be one prior paper published about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
different from this parser: it attempts to present correlated log data in a
form suitable for a systems administrator to search using the myriad of
standard Unix text processing utilities already available, producing both a
\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
support both \CSV{} and Berkeley DB appears to have been a limiting factor:
\CSV{} is a very simple format where a record is stored in a single line,
with fields separated by a comma or other punctuation symbol.  Problems
with \CSV{} files include the difficulty in escaping separators, providing
multiple values for a field (e.g.\ multiple recipients) and adding new
fields.  There is no facility to include documentation of the fields,
separator, etc., unlike \SQL{} databases where every database includes a
schema naming the fields and the type of data they store (integer, text,
timestamp, etc.).

>>>>

XXX CLARIFY THE SENTENCE STARTING WITH ``Although this may see'' IN THE
NEXT PARAGRAPH\@.

@@

XXX CHANGE THIS\@; THE PARSER DESIGN IS NOVEL, DO NOT STRESS THE SIMILARITY
TOO MUCH\@.

@@

There only appears to be one prior paper published about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
different from this parser: it attempts to present correlated log data in a
form suitable for a systems administrator to search using the myriad of
standard Unix text processing utilities already available, producing both a
\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
support both \CSV{} and Berkeley DB appears to have been a limiting factor:
\CSV{} is a very simple format where a record is stored in a single line,
with fields separated by a comma or other punctuation symbol.  Problems
with \CSV{} files include the difficulty in escaping separators, providing
multiple values for a field (e.g.\ multiple recipients) and adding new
fields.  There is no facility to document the fields or the separator,
unlike \SQL{} databases where every database includes a schema naming the
fields and the type of data they store (integer, text, timestamp, etc.).

### 2008-07-09 doc/logparser-coverage.tex
Corrections from Mads and chktex.

Warnings are issued for any log lines which are not parsed; no warnings are
issued for unparsed log lines while testing with the \numberOFlogFILES{}
test log files, so it can be safely concluded that there are zero false
negatives.  False positives are harder to quantify: short of examining each
of the 60,721,709 log lines and determining which \regex{} parsed it, there
is no way to be sure that every line was parsed by the correct \regex{},
making it impossible to quantify the false positive rate; however a random
sample of 6039 log lines was parsed and the results checked manually to
ensure that the correct \regex{} parsed each log line.\footnote{Each log
line was examined and the correct \regex{} identified from the
\numberOFrules{} rules in the database; the correct \regex{} was then
compared to the \regex{} which was used by the parser.}  The sample was
generated by running the following command:

>>>>

Warnings are issued for any log lines which are not parsed; no warnings are
issued for unparsed log lines while testing with the \numberOFlogFILES{}
test log files, so it can be safely concluded that there are zero false
negatives.  False positives are harder to quantify: short of examining each
of the 60,721,709 log lines and determining which \regex{} parsed it, there
is no way to be sure that every line was parsed by the correct \regex{},
making it impossible to quantify the false positive rate; however a random
sample of 6,039 log lines was parsed and the results checked manually to
ensure that the correct \regex{} parsed each log line.\footnote{Each log
line was examined and the correct \regex{} identified from the
\numberOFrules{} rules in the database; the correct \regex{} was then
compared to the \regex{} which was used by the parser.}  The sample was
generated by running the following command:

### 2008-07-09 doc/logparser-glossary.tex
Corrections from Mads and chktex.

% The contents of the glossary.
\glossary{name={SQLite3},description={
    \textit{SQLite is a small C library that implements a self-contained,
    embeddable, zero-configuration SQL database engine.\/} SQLite3 is an
    \SQL{} implementation focusing on correctness, simplicity and speed.
    Unlike other \SQL{} implementations it does not require a separate
    server process, greatly simplifying deployment of an application
    utilising it.  More details can be found at~\cite{sqlite-features} or
    \url{http://www.sqlite.org/}.
}}

@@

\glossary{name={awk},description={AWK is a general purpose programming
    language that is designed for processing text-based data, and is
    available as a standard utility on all Unix systems.
}}

\glossary{name={syslog},description={Syslog is the standard logging
    mechanism on Unix systems: the program sends log messages to syslog,
    then syslog filters and stores the messages according to the
    configuration the administrator has chosen.
}}

>>>>

% vim: set filetype=tex :
% The contents of the glossary.
\glossary{name={SQLite3},description={
    \textit{SQLite is a small C library that implements a self-contained,
    embeddable, zero-configuration SQL database engine.\/} SQLite3 is an
    \SQL{} implementation focusing on correctness, simplicity and speed.
    Unlike other \SQL{} implementations it does not require a separate
    server process, greatly simplifying deployment of an application
    utilising it.  More details can be found at~\cite{sqlite-features} or
    \url{http://www.sqlite.org/}.
}}

@@

\glossary{name={awk},description={
    AWK is a general purpose programming language that is designed for
    processing text-based data, and is available as a standard utility on
    all Unix systems.
}}

\glossary{name={syslog},description={
    Syslog is the standard logging mechanism on Unix systems: the program
    sends log messages to syslog, then syslog filters and stores the
    messages according to the configuration the administrator has chosen.
}}

\glossary{name={mail bomb},description={
    A mail bomb occurs when an attacker inflicts a huge volume of mail on
    the victim.  At best a mail bomb is irritating to the victim; at worst
    the deluge of mail can be severe enough to: interrupt service for the
    victim and/or other users; cause mail to be rejected because the victim
    has reached a limit (e.g.\ too many mails, too much disk space
    consumed); the victim may accidentally delete other mail while trying
    to cope with the mail bomb.
}}

\glossary{name={mail loop},description={
    Sometimes mail set to one address must be delivered to a different
    address instead, e.g.\ because a person has changed jobs.  A mail loop
    occurs when the recipient addresses constitute a cyclic directed graph;
    the simplest example is when mail for \texttt{foo\at{}example.net} is
    delivered to \texttt{bar\at{}example.com}, and mail for
    \texttt{bar\at{}example.com} is delivered to
    \texttt{bar\at{}example.com}.
}}

### 2008-07-09 doc/logparser-other-parsers-review.tex
Corrections from Mads and chktex.

\subsection{Parsers reviewed}

In this section text in italics is a quote extracted from the documentation
of the program being reviewed.

\subsubsection{Pflogsumm}

\textit{pflogsumm is designed to provide an over-view of Postfix activity,
with just enough detail to give the administrator a ``heads up'' for
potential trouble spots.\/}

@@

\subsubsection{Sawmill Universal Log File Analysis and Reporting}

\textit{ Sawmill is a Postfix log analyzer (it also support 686 other log
formats).  It can process log files in Postfix format, and generate dynamic
statistics from them, analyzing and reporting events.  Sawmill can parse
Postfix logs, import them into a SQL database (or its own built-in
database), aggregate them, and generate dynamically filtered reports, all
through a web interface.  Sawmill can perform Postfix analysis on any
platform, including Window, Linux, FreeBSD, OpenBSD, Mac OS, Solaris, other
UNIX, and more.\/}

@@

\subsubsection{Splunk}

\textit{Splunk is an IT Search engine. It is software that indexes any
format of IT data from any source in real time, including logs,
configurations, scripts, code, messages, traps, alerts, activity reports,
stack traces and metrics from all of your applications, servers and
devices. Splunk lets you search, navigate, alert and report on all your IT
data in real time using an AJAX web interface. You can also share knowledge
and Splunk solutions with other members of the Splunk community via
SplunkBase.\/}

@@

Splunk supports reading compressed log files without any configuration by
the user.  The free version of Splunk limits the volume of data indexed per
day to 500MB, though a trial Enterprise licence is available which allows
indexing of up to 5GB of data per day.  The cheapest licensed version costs
\$5000 (plus \$1000 support), and still limits the volume of data indexed
per day to 500MB.

@@

\subsubsection{Isoqlog}

\textit{Isoqlog is an MTA log analysis program written in C. It designed to
scan qmail, postfix, sendmail and exim logfile and produce usage statistics
in \HTML{} format for viewing through a browser. It produces Top domains
output according to Sender, Receiver, Total mails and bytes; it keeps your
main domain mail statistics with regard to Days Top Domain, Top Users
values for per day, per month and years.\/}

@@

\subsubsection{AWStats}

\textit{AWStats is a free powerful and featureful tool that generates
advanced web, streaming, ftp or mail server statistics, graphically. This
log analyzer works as a CGI or from command line and shows you all possible
information your log contains, in few graphical web pages. It uses a
partial information file to be able to process large log files, often and
quickly. It can analyze log files from all major server tools like Apache
log files (NCSA combined/XLF/ELF log format or common/CLF log format),
WebStar, IIS (W3C log format) and a lot of other web, proxy, wap, streaming
servers, mail servers and some ftp servers.\/}

@@

\subsubsection{Throughput monitor}

@@

\subsubsection{Anteater}

\textit{The Anteater project is a Mail Traffic Analyser. Anteater supports
currently the logformat produced by Sendmail and by Postfix. The tool is
written in 100\% C++ and is very easy to customize. Input, output, and the
analysis are modular class objects with a clear interface. There are eight
useful analyse modules, writing the result in plain ASCII or \HTML{}, to
stdout or to files.\/}

@@

\subsubsection{Yet Another Advanced Logfile Analyser}

\textit{yaala is a very flexible analyser for all kinds of logfiles. It
uses parsers to extract information from a logfile, an SQL-like query
language to relate the information to each other and an output-module to
format the information appropriately.\/}

@@

\subsubsection{Lire}

\textit{As any good system administrator knows, there's a lot more to keep
track of in an active network than just webservers. Lire is hands down the
most versatile log analysis software available today. Lire not only keeps
you informed about your HTTP, FTP, and mail traffic, it also reports on
your firewalls, your print servers, and your DNS activity. The ever growing
list of Lire-supported services clearly outstrips any other software, in
large part thanks to the numerous volunteers who have pioneered many new
services and features. Lire is a total solution for your log analysis
needs.\/}

@@

\subsubsection{Logrep}

\textit{Logrep is a secure multi-platform framework for the collection,
extraction, and presentation of information from various log files. It
features HTML reports, multi dimensional analysis, overview pages, SSH
communication, and graphs, and supports over 30 popular systems including
Snort, Squid, Postfix, Apache, Sendmail, syslog, ipchains, iptables, NT
event logs, Firewall-1, wtmp, xferlog, Oracle listener and Pix.\/}

@@

\subsubsection{Log Mail Analyser}

@@

\begin{table}[ht]
\begin{tabular}{l|l|l|l|l|l}
    Parser         & Parsed test & Data        & Custom        & Documentation  & Source \\
                   & log files?  & store?      & reports?      & quality?       & code?  \\
    \hline         &             &             &               &                &        \\
    Pflogsumm      & Yes         & No          & Partial\dag   & Good           & Yes    \\
    Sawmill        & Yes         & Yes         & Searches      & Reasonable     & No     \\
    Splunk         & Yes         & Yes         & Searches      & Abundant       & No     \\
                   &             &             & \& reports    & but poor       &        \\
    Isoqlog        & No          & Yes         & No            & No             & Yes    \\
    AWStats        & Partially   & Yes         & Partial\dag   & Good           & Yes    \\
    Throughput     & N/A         & No          & No            & Brief but good & Yes    \\
    \tab{}monitor  &             &             &               &                &        \\
    Anteater       & No          & No          & No            & No             & Yes    \\
    YAALA          & No          & Yes\ddag    & Searches      & Poor           & Yes    \\
    Lire           & Yes         & Yes         & Yes           & Yes            & Yes    \\
    Logrep         & No          & Yes         & No            & Barely         & Yes    \\
    \LMA{}         & No          & Yes         & No            & No             & Yes    \\ 
    \parsername{}  & Yes         & Yes\dag\dag & No            & Yes            & Yes    \\
\end{tabular}

\dag Sections can be omitted from a report, but extra sections can not be
added.

\ddag YAALA only stores the data required to produce the latest report;
other data will be discarded.

\dag\dag\parsername{} is the only parser with documentation for its data
store.

>>>>

\subsection{Pflogsumm}

\begin{quotation}

    pflogsumm is designed to provide an over-view of Postfix activity, with
    just enough detail to give the administrator a ``heads up'' for
    potential trouble spots.

\end{quotation}

@@

\subsection{Sawmill Universal Log File Analysis and Reporting}

\begin{quotation}

    Sawmill is a Postfix log analyzer (it also support [sic] 686 other log
    formats).  It can process log files in Postfix format, and generate
    dynamic statistics from them, analyzing and reporting events.  Sawmill
    can parse Postfix logs, import them into a SQL database (or its own
    built-in database), aggregate them, and generate dynamically filtered
    reports, all through a web interface.  Sawmill can perform Postfix
    analysis on any platform, including Window, Linux, FreeBSD, OpenBSD,
    Mac OS, Solaris, other UNIX, and more.

\end{quotation}

@@

\subsection{Splunk}

\begin{quotation}

    Splunk is an IT Search engine. It is software that indexes any format
    of IT data from any source in real time, including logs,
    configurations, scripts, code, messages, traps, alerts, activity
    reports, stack traces and metrics from all of your applications,
    servers and devices. Splunk lets you search, navigate, alert and report
    on all your IT data in real time using an AJAX web interface. You can
    also share knowledge and Splunk solutions with other members of the
    Splunk community via SplunkBase.

\end{quotation}

@@

Splunk supports reading compressed log files without any configuration by
the user.  The free version of Splunk limits the volume of data indexed per
day to 500MB, though a trial Enterprise licence is available which allows
indexing of up to 5GB of data per day.  The cheapest licensed version costs
\$5000 (plus \$1000 support), and still limits the volume of data indexed
per day to 500MB\@.

@@

\subsection{Isoqlog}

\begin{quotation}

    Isoqlog is an MTA log analysis program written in C. It designed to
    scan qmail, postfix, sendmail and exim logfile and produce usage
    statistics in \HTML{} format for viewing through a browser. It produces
    Top domains output according to Sender, Receiver, Total mails and
    bytes; it keeps your main domain mail statistics with regard to Days
    Top Domain, Top Users values for per day, per month and years.

\end{quotation}

@@

\subsection{AWStats}

\begin{quotation}

    AWStats is a free powerful and featureful tool that generates advanced
    web, streaming, ftp or mail server statistics, graphically. This log
    analyzer works as a CGI or from command line and shows you all possible
    information your log contains, in few graphical web pages. It uses a
    partial information file to be able to process large log files, often
    and quickly. It can analyze log files from all major server tools like
    Apache log files (NCSA combined/XLF/ELF log format or common/CLF log
    format), WebStar, IIS (W3C log format) and a lot of other web, proxy,
    wap, streaming servers, mail servers and some ftp servers.

\end{quotation}

@@

\subsection{Throughput monitor}

@@

\subsection{Anteater}

\begin{quotation}

    The Anteater project is a Mail Traffic Analyser. Anteater supports
    currently the logformat produced by Sendmail and by Postfix. The tool
    is written in 100\% C++ and is very easy to customize. Input, output,
    and the analysis are modular class objects with a clear interface.
    There are eight useful analyse modules, writing the result in plain
    ASCII or \HTML{}, to stdout or to files.

\end{quotation}

@@

\subsection{Yet Another Advanced Logfile Analyser}

\begin{quotation}

    yaala is a very flexible analyser for all kinds of logfiles. It uses
    parsers to extract information from a logfile, an SQL-like query
    language to relate the information to each other and an output-module
    to format the information appropriately.

\end{quotation}

@@

\subsection{Lire}

\begin{quotation}

    As any good system administrator knows, there's a lot more to keep
    track of in an active network than just webservers. Lire is hands down
    the most versatile log analysis software available today. Lire not only
    keeps you informed about your HTTP, FTP, and mail traffic, it also
    reports on your firewalls, your print servers, and your DNS activity.
    The ever growing list of Lire-supported services clearly outstrips any
    other software, in large part thanks to the numerous volunteers who
    have pioneered many new services and features. Lire is a total solution
    for your log analysis needs.

\end{quotation}

@@

\subsection{Logrep}

\begin{quotation}

    Logrep is a secure multi-platform framework for the collection,
    extraction, and presentation of information from various log files. It
    features HTML reports, multi dimensional analysis, overview pages, SSH
    communication, and graphs, and supports over 30 popular systems
    including Snort, Squid, Postfix, Apache, Sendmail, syslog, ipchains,
    iptables, NT event logs, Firewall-1, wtmp, xferlog, Oracle listener and
    Pix.

\end{quotation}

@@

\subsection{Log Mail Analyser}

@@

\begin{table}[ht]
\begin{tabular}{l|l|l|l|l|l}
    Parser         & Parsed test & Data            & Custom        & Documentation  & Source \\
                   & log files?  & store?          & reports?      & quality?       & code?  \\
    \hline         &             &                 &               &                &        \\
    Pflogsumm      & Yes         & No              & Partial\dag{} & Good           & Yes    \\
    Sawmill        & Yes         & Yes             & Searches      & Reasonable     & No     \\
    Splunk         & Yes         & Yes             & Searches      & Abundant       & No     \\
                   &             &                 & \& reports    & but poor       &        \\
    Isoqlog        & No          & Yes             & No            & No             & Yes    \\
    AWStats        & Partially   & Yes             & Partial\dag{} & Good           & Yes    \\
    Throughput     & N/A         & No              & No            & Brief but good & Yes    \\
    \tab{}monitor  &             &                 &               &                &        \\
    Anteater       & No          & No              & No            & No             & Yes    \\
    YAALA          & No          & Yes\ddag{}      & Searches      & Poor           & Yes    \\
    Lire           & Yes         & Yes             & Yes           & Yes            & Yes    \\
    Logrep         & No          & Yes             & No            & Barely         & Yes    \\
    \LMA{}         & No          & Yes             & No            & No             & Yes    \\ 
    \parsername{}  & Yes         & Yes\dag{}\dag{} & No            & Yes            & Yes    \\
\end{tabular}

\dag{} Sections can be omitted from a report, but extra sections can not be
added.

\ddag{} YAALA only stores the data required to produce the latest report;
other data will be discarded.

\dag{}\dag{}\parsername{} is the only parser with documentation for its
data store.

### 2008-07-09 doc/logparser-parsing-algorithm.tex
Corrections from Mads and chktex.

@@

Because of this the parser cannot complain when it encounters a log line
from \daemon{qmgr} for a previously unseen mail; it must flag the mail as
coming from an unknown origin, and subsequently clear the flag if and when
the origin of the mail becomes clear.  Obviously the parser could omit
checking of where mails originate from, but the author believes that it is
better to require an explicit source, as bugs in the parser are more likely
to be exposed; such checks helped to identify the complications described
in \sectionref{discarding cleanup lines} and \sectionref{pickup logging
after cleanup}.

@@

Figures~\refwithpage{flow chart image part 1} and~\refwithpage{flow chart
image part 2} show the paths the data representing a mail can take through
the parser algorithm.  The flow chart covers the most common paths only;
there are additional, uncommon paths which are excluded for the sake of
clarity; details of the deviations can be found in \sectionref{additional
complications}.

>>>>

XXX DUNNO IF THIS SECTION SHOULD BE HERE\@.  QUEUEID VS PID ISN'T TOO
IMPORTANT\@; CONNECTION REUSE COULD EASILY BE MOVED TO ADDITIONAL
COMPLICATIONS\@; RE-INJECTED MAILS SHOULD PROBABLY BE MOVED TO ADDITIONAL
COMPLICATIONS TOO\@.

@@

Because of this the parser cannot complain when it encounters a log line
from \daemon{qmgr} for a previously unseen mail; it must flag the mail as
coming from an unknown origin, and subsequently clear the flag if and when
the origin of the mail becomes clear.  Obviously the parser could omit
checking of where mails originate from, but requiring an explicit source
helps to expose bugs in the parser; such checks helped to identify the
complications described in \sectionref{discarding cleanup lines} and
\sectionref{pickup logging after cleanup}.

@@

XXX CONSIDER REMOVING PART TWO OF THE FLOW CHART\@: I DON'T THINK THAT
RE-INJECTION IS IMPORTANT ENOUGH TO WARRANT IT\@.  THE EXPLANATION OF
TRACKING (\textsection{}5.4.3) COULD BE MOVED TO ADDITIONAL
COMPLICATIONS\@.

Figure~\refwithpage{flow chart image part 1} and figure~\refwithpage{flow
chart image part 2} show the paths the data representing a mail can take
through the parser algorithm.  The flow chart covers the most common paths
only; there are additional, uncommon paths which are excluded for the sake
of clarity; details of the deviations can be found in
\sectionref{additional complications}.
